---
title: "Simulation using ComplexSim"
author: " "
date: '`r format(Sys.time(), "%d %B, %Y")`'
params:
  Sample_size:
    label: "Simulated sample size (number of patients)"
    value: 50
  N_Genes:
    label: "Simulated number of genes"
    value: 100
  Time_points:
    label: "Number of time points when repeated measures are taken"
    value: 5
  Group_total:
    label: "Number of groups"
    value: 3
  Control_group_total:
    label: "Number of control groups"
    value: 1
  Treatment_group_total:
    label: "Number of treatment groups"
    value: 2
  Multi_treatment:
    label: "Each sample takes all the treatments and controls in independent environments"
    value: "Y"
  eps_dependent:
    label: "Proportion of dependent subjects in different groups, should be 1 by default if Multi_treatment = Y. "
    value: 0.2
  eps_SignalGenes: 
    label: "Proportion of true signal genes"
    value: 0.2
  beta0:
    label: "Average effect"
    value: 0
  beta1:
    label: "Treatment group effect"
    value: 4
  beta2:
    label: "Time effect"
    value: 2
  mu:
    label: "Mean of the Gaussian vector defining random intercept and slope"
    value: [0,0]
  sigma:
    lable: "Variance of the Gaussian vector defining random intercept and slope"
    value: 1.5
  tau:
    label: "Variance-covariance elements of the random intercept and slope"
    value: [1,1,0]
  act.cor:
    value: 0.6
  Sub_xy:
    label: "Subject-level correlation defined on the pairs of subjects."
    value: !expr list(c(1,2), c(2,3), c(1,3), c(4,7))
  Sub_corr:
    lable: "Pairwise subject-level correlation coefficient."
    value: [0.8, 0.6, 0.3, 0.8]
  sigma_RE:
    label: "Standard deviation of the random block factor."
    value: 0.8
  Rep_corr:
    label: ""
    value: 0.6
  Balanced_visit:
    label: "Each sample has same number of visits"
    value: "Y"
  Fixed_form:
    label: "Basic linear model with fixed effects"
    value: "Exprs ~ 0 + GROUP*VISIT"
  Mixed_form:
    label: "Mixed-effect model"
    value: "Exprs ~ GROUP + VISIT + (1|ID)"
  RunPkg_dependence:
    value: TRUE
  Run_parallel:
    value: TRUE
editor_options: 
  chunk_output_type: console
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Simulation}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
---

```{r setup, include=FALSE}
# knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(eval = T, echo = TRUE, results = 'asis', 
                      message = F, warning = F, error = T,
                      width = 120)
```


```{r packages}
library(MASS)
library(utils)
library(nlme)
library(edgeR)
library(parallel) ##
library(RhpcBLASctl) ##
# library(CovStructRM) ## 
library(emmeans)
library(reshape2)
library(tidyverse)
# library(ComplexSim)

source("../R/sim_all.R")

theme_set(theme_bw())
```

# Introduction

When we have group effect, repeated measures and subject level variability in the experiment, we usually define a mixed effect linear model as 
\begin{equation}
y_{ijk} = \beta_0 + \beta_i + \beta_j + \alpha_k + \lambda_{ij} + \varepsilon_{ijk}, 
\end{equation}
where the individual-level random effect $\alpha_k$ can be regarded as `random intercept`. 

In order to include both random intercept and random slope, we should consider the model

\begin{equation}
y_{ijk} = \left( \beta_0 + \alpha_k^{(0)} \right)+ \left(\beta_1 + \alpha_k^{(1)} \right) x_{ijk}  + \varepsilon_{ijk}. 
\end{equation}

The goal of this vignette is to provide several examples of different designs and a comparison of the performance of different approaches to fit the models.


# The design 1

In the first example we provide the simulation with the following design:

* Each individual has samples in multiple groups
* With repeated measures

## Define the design

The function `wrap_design` gives the desired model structure. One can define a data matrix by inserting 

* `n` = number of patients in each group,  
* `m` = number of repeated visits (time points),
* `q` = number of groups (eg., control and multiple treatment groups)
* `SingleGroup = FALSE` indicates that each patient has samples in multiple groups.

```{r Simulateion design 1, results='markup'}
#############################################################
###
### TEST 
###
#############################################################
dat <- wrap_design(n = params$Sample_size, 
                   m = params$Time_points, 
                   q = params$Group_total,
                   SingleGroup = FALSE, 
                   eps = FALSE)

cat("\n Sample size per group per time point: \n")

dat %>% distinct(ID, GROUP, VISIT) %>% 
  dplyr::count(GROUP, VISIT) %>% knitr::kable()

dat[1:10, ] %>%  knitr::kable()
```

## Create simulated data based on Gaussian distribution

The function `wrap_simulation` uses the model structure as input, and generates the simulated data set in a long format data frame.

* `data` = Design matrix in long format. 
* $\beta_0$ = Mean of the global effect.
* $\beta_1$ = Mean of group effect.
* $\beta_2$ = Mean of time effect. 
* $\Lambda$ = Coefficient of interaction terms. 
* $\sigma$ = Standard deviation of the simulated AR(1) sequence.
* $\mu$ = Mean vector of random intercept and slope. 
* $\tau$ = Elements of variance-covariance matrix of random intercept and slope.
* `act.cor` = Correlation parameter of the AR(1) structure.

```{r, results='markup'}
dat <- wrap_simulation(data = dat, 
                       SingleGroup = FALSE, 
                       beta0 = params$beta0,
                       beta1 = params$beta1*(1:params$Group_total),
                       beta2 = log((params$beta2*(1:params$Time_points+1))),
                       lambda = NULL,
                       sigma = params$sigma,
                       mu = params$mu,
                       act.cor = params$act.cor,
                       tau = params$tau) 

dat$GeneName <- "GENE_1"
dat[1:10, ] %>%  knitr::kable()
```

Plot profile

```{r}
### Plot profile
plot_prof(dat)
```

Empirical correlation

```{r Empirical correlation design 1, results='markup', eval=FALSE}
### Corr
# CovCor_err(dat, condition = "Group2")
# 
# CovCor_err(dat, condition = "Group3")

### Plot corr
CovCor_err(dat, condition = "Group2", plot.line = TRUE)
CovCor_err(dat, condition = "Group2", plot.heat = TRUE)

CovCor_err(dat, condition = "Group3", plot.line = TRUE)
CovCor_err(dat, condition = "Group3", plot.heat = TRUE)
```

## Create simulated data based on Negative Binomial distribution

```{r}
dat <- wrap_design(n = params$Sample_size, 
                   m = params$Time_points, 
                   q = params$Group_total,
                   SingleGroup = FALSE, 
                   eps = FALSE)

dat <- wrap_simulation_NB(dat, 
                             SingleGroup = FALSE, 
                             tau = c(1,1,0)) ### NB
# 
dat$GeneName <- "GENE_1"
dat[1:10, ] %>%  knitr::kable()
### Plot profile
plot_prof(dat)
```


# The design 2

## Create simulated data based on Gaussian distribution

```{r Simulate design 2}
#############################################################
###
### TEST 
###
#############################################################
dat <- wrap_design(n = params$Sample_size, 
                   m = params$Time_points, 
                   q = params$Group_total, 
                   SingleGroup = TRUE,
                   eps = FALSE)
### 
dat <- wrap_simulation(dat, 
                       SingleGroup = TRUE, 
                       beta0 = params$beta0,
                       beta1 = params$beta1*(1:params$Group_total),
                       beta2 = log((params$beta2*(1:params$Time_points+1))),
                       lambda = NULL,
                       sigma = params$sigma,
                       mu = params$mu,
                       tau = params$tau,
                       act.cor = params$act.cor) ### Variables
dat$GeneName <- "GENE_1"
# dat[1:10, ] %>%  knitr::kable()
# dat[90:100, ] %>%  knitr::kable()
```

Plot profile

```{r}
### Plot profile
plot_prof(dat)
```

Empirical correlation

```{r, results='markup', eval=FALSE}
### Corr
# CovCor_err(dat, condition = "Group2")
# CovCor_err(dat, condition = "Group3")

### Plot corr
CovCor_err(dat, condition = "Group2", plot.line = TRUE)
CovCor_err(dat, condition = "Group2", plot.heat = TRUE)

CovCor_err(dat, condition = "Group3", plot.line = TRUE)
CovCor_err(dat, condition = "Group3", plot.heat = TRUE)
```

## Create simulated data based on Negative Binomial distribution


```{r}
dat <- wrap_design(n = params$Sample_size, 
                   m = params$Time_points, 
                   q = params$Group_total, 
                   SingleGroup = TRUE,
                   eps = FALSE)
dat <- wrap_simulation_NB(dat, SingleGroup = TRUE, tau = c(1,1,0)) ### NB
dat$GeneName <- "GENE_1"
### Plot profile
plot_prof(dat)
```


# The design 3

For example, we consider the following scenario:

* Group 1: P1, P2, P3, P4, P5, P6, P7, P8, P9, P10
* Group 2: P1, P2, P13, P14, P15, P16, P17, P18, P19, P20
* Group 3: P1, P2, P23, P24, P25, P26, P27, P28, P29, P30

Suppose P1 and P2 appear repeatedly in Group 1, 2 and 3. 
They can also be regarded as highly correlated samples instead of identical ones in different groups. 

**Based on design 2, we now have `SingleGroup=TRUE`, sample size in each group `n=10` and the ratio of identical sample `eps=0.2`.** 
The number of repeated subjects is $n\varepsilon = 2$. 
All the other subjects are independent. 

Let $y_{ijk}$ be the normalized data of individual $k$ in group $i$, at time point $j$. With the relationship between $i$ and $k$, we have the following setting: 
The total number of subjects in each group are identical: $n_1 = n_2 = \cdots = n.$ 
There are repeated subjects nRep and independent subjects (n - nRep) in each group. 


## Create simulated data based on Gaussian distribution

```{r Simulate design 3}
#############################################################
###
### TEST 
###
#############################################################
dat <- wrap_design(n = params$Sample_size, 
                   m = params$Time_points, 
                   q = params$Group_total, 
                   SingleGroup = TRUE, 
                   eps = params$eps_dependent)
### 
dat <- wrap_simulation(dat, 
                       SingleGroup = TRUE, 
                       beta0 = params$beta0,
                       beta1 = params$beta1*(1:params$Group_total),
                       beta2 = log((params$beta2*(1:params$Time_points+1))),
                       lambda = NULL,
                       sigma = params$sigma,
                       mu = params$mu,
                       tau = params$tau,
                       act.cor = params$act.cor) ### Variables
dat$GeneName <- "GENE_1"
# dat[1:10, ] %>%  knitr::kable()
# dat[90:100, ] %>%  knitr::kable()
```

Plot profile

```{r}
### Plot profile
plot_prof(dat)
```

Empirical correlation

```{r, results='markup', eval=FALSE}
# ### Corr
# CovCor_err(dat, condition = "Group2")
# 
# CovCor_err(dat, condition = "Group3")

### Plot corr
CovCor_err(dat, condition = "Group2", plot.line = TRUE)
CovCor_err(dat, condition = "Group2", plot.heat = TRUE)

CovCor_err(dat, condition = "Group3", plot.line = TRUE)
CovCor_err(dat, condition = "Group3", plot.heat = TRUE)
```


## Create simulated data based on Negative Binomial distribution


```{r}
dat <- wrap_design(n = params$Sample_size, 
                   m = params$Time_points, 
                   q = params$Group_total, 
                   SingleGroup = TRUE, 
                   eps = params$eps_dependent)
dat <- wrap_simulation_NB(dat, SingleGroup = TRUE, tau = c(1,1,0)) ### NB
dat$GeneName <- "GENE_1"
### Plot profile
plot_prof(dat)
```


# Simulate i.i.d. gene set 

Now we use function `wrap_sim_DGE` to simulate the expression data for more than one genes, and convert the readouts to a DGEList. 

```{r Create iid DGEList, results='markup', warning=FALSE}
trt_genewise <- matrix(seq(1:(params$Group_total*params$N_Genes)),
                       ncol=params$N_Genes)

sim_DGEList <- wrap_sim_DGEList(n=params$Sample_size, 
                                m=params$Time_points, 
                                q=params$Group_total, 
                                params$N_Genes, 
                                trt_eff = trt_genewise,
                                distribution = NULL,
                                SingleGroup=FALSE)
# str(sim_DGEList)
```


```{r, results='markup'}
head(sim_DGEList$norm_counts)[, 1:10]
head(sim_DGEList$pheno_data)
```


```{r, results='markup', warning=FALSE, eval=FALSE}
trt_genewise <- matrix(seq(1:(params$Group_total*params$N_Genes)),
                       ncol=params$N_Genes)

sim_DGEList <- wrap_sim_DGEList(n=params$Sample_size, 
                                m=params$Time_points, 
                                q=params$Group_total, 
                                params$N_Genes, 
                                trt_eff = trt_genewise, 
                                distribution = "NB",
                                SingleGroup=FALSE)
# str(sim_DGEList)
```


```{r, results='markup'}
head(sim_DGEList$norm_counts)[, 1:10]
head(sim_DGEList$pheno_data)
```

# Simulate mixture of genes with user-defined structure 


With default params:

```{r Simulate default mixture genes, results='markup'}
dat <- wrap_design(n = params$Sample_size, 
                   m = params$Time_points, 
                   q = params$Group_total, 
                   SingleGroup = TRUE)
# view(dat)
### 
# dat <- wrap_simulation(dat, SingleGroup = TRUE, tau = c(1,1,0)) ### Variables
sim_DGEList <- wrap_sim_corr(dat, 
                             N = params$N_Genes, 
                             eps_Signal = params$eps_SignalGenes, 
                             SingleGroup = TRUE, 
                             act.cor = NULL)

# head(sim_DGEList$norm_counts, 1)

DT::datatable(sim_DGEList$pheno_data)
```


# With user-defined params

<!-- ## TO DO: parametrize the functions used in generating the group mean vector per gene. Is there an easy way to do it? (ZJ) -->

```{r Simulate arbitrary gene set, results='markup'}
dat <- wrap_design(n = params$Sample_size, 
                   m = params$Time_points, 
                   q = params$Group_total, 
                   SingleGroup = TRUE)
# dat <- wrap_simulation(dat, SingleGroup = TRUE, tau = c(1,1,0)) ### Variables
G_Signal <- ceiling(params$N_Genes*params$eps_SignalGenes)
G_Null <- params$N_Genes - G_Signal 

coef.beta2_Signal <- lapply(1:G_Signal, 
                            function(x) {seq(1, params$Time_points, length = params$Time_points)})
coef.beta2_null <- lapply(1:G_Null, 
                                function(x) {rep(0, params$Time_points)})
Time_effect <- c(coef.beta2_Signal, coef.beta2_null)


coef.beta1_Signal <- lapply(1:G_Signal, 
                                  function(x) {seq(1, 2*params$Group_total, length = params$Group_total)})
coef.beta1_null <- lapply(1:G_Null, 
                                function(x) {rep(0, params$Group_total)})
TRT_effect <- c(coef.beta1_Signal, coef.beta1_null)

###
sim_DGEList <- wrap_sim_corr(dat, 
                             params$N_Genes, 
                             params$eps_SignalGenes, 
                             SingleGroup = TRUE,
                             coef.beta2 = Time_effect, 
                             act.cor = NULL)

head(sim_DGEList$norm_counts, 1)

# DT::datatable(sim_DGEList$pheno_data)
```

## Dependent subjects

One can simulate arbitrary gene sets with user defined correlated subjects.

<!-- ## TO DO: Modify the elements according to the list given in yaml params. -->

```{r Simulate arbitrary gene set with dependent subjects, results='markup'}
### test the simulation with covariance
corr_Sub = diag(params$Sample_size*params$Group_total) 

for (i in 1:length(params$Sub_corr)) {
corr_Sub[params$Sub_xy[[i]][1], params$Sub_xy[[i]][2]] = params$Sub_corr[i]^2
corr_Sub[params$Sub_xy[[i]][2], params$Sub_xy[[i]][1]] = params$Sub_corr[i]^2
}
CovHeatPlot(corr_Sub[1:params$Sample_size, 1:params$Sample_size])
CovHeatPlot(corr_Sub[1:10, 1:10])
sigma_Group = c(1:params$Group_total)

corr_Rep = fVarAR1(0.6, params$Time_points)
sigma_Rep = seq(1, 2, length=params$Time_points)

SimCov <- CovMarginal_sim(corr_Sub, sigma_Group, corr_Rep, sigma_Rep) 

DT::datatable(SimCov[1:params$Sample_size, 1:params$Sample_size])
# View(SimCov)
# ComplexHeatmap::pheatmap(corr_Sub, cluster_rows = FALSE, cluster_cols = FALSE)

CovHeatPlot(SimCov[1:params$Sample_size, 1:params$Sample_size])
# CovHeatPlot(SimCov[1:10, 1:10])

# View(SimCov[c(1:10, 41:50), c(1:10, 41:50)])
# CovHeatPlot(SimCov[c(1:10, 41:50), c(1:10, 41:50)])

sim_DGEList <- wrap_sim_corr(dat, 
                             params$N_Genes,
                             params$eps_SignalGenes, 
                             SingleGroup = TRUE,
                             coef.beta1 = TRT_effect,
                             coef.beta2 = Time_effect, 
                             CovMarginal = SimCov,
                             act.cor = NULL)

str(sim_DGEList)
```


```{r, eval=FALSE}
# save simulated data to file
filename = paste0(getwd(), "/data/Simulation_ArbitraryFamily.csv")

norm_counts_wider <- t(sim_DGEList$norm_counts) %>% as.data.frame() %>%
  rownames_to_column(., var = "SampleID") 
annot_table <- sim_DGEList$pheno_data %>% as.data.frame() %>%
  rownames_to_column(., var = "SampleID") %>%
  mutate(FAMILY = ID)
ListFamily <- lapply(1:length(params$Sub_xy), function(n){paste0("P", params$Sub_xy[[n]])} ) 
unlistFamily <- data.frame()
for (FamilyNum in 1:length(ListFamily)) {
  aFamily <- data.frame(SUBID = ListFamily[[FamilyNum]],
               FAMILY = paste0("Family_", FamilyNum))
  unlistFamily <- rbind(unlistFamily, aFamily) 
}
FamilyCluster <- unlistFamily %>% group_by(SUBID) %>% dplyr::summarise(UniqueFamily = n()) %>% filter(UniqueFamily > 1) %>% pull(SUBID)
for (PID in FamilyCluster) {
  unlistFamily_Unique <- unlistFamily %>% 
    filter(ID %in% PID)
}
# for (FamilyNum in unlistFamily$FAMILY ) {
# }
for (FamilyNum in 1:length(ListFamily)) {
  annot_table <- annot_table %>% mutate(
    FAMILY = ifelse(ID %in% ListFamily[[FamilyNum]], paste0("Family_", FamilyNum), FAMILY)
  )
}
norm_counts_wider <- annot_table %>% left_join( norm_counts_wider, by=c("SampleID"))

norm_counts_long <- norm_counts_wider %>% tidyr::pivot_longer(
  cols=!(c("SampleID", "ID", "FAMILY",  "VISIT", "GROUP")),
  names_to = "GeneName", 
  values_to = "Exprs")

sim_DGE_long_ArbitraryFamily <- norm_counts_long

write.csv(sim_DGE_long_ArbitraryFamily, file = filename)
```

## Twins

```{r Simulate_TWINS}
params$Sub_corr <- rep(0.95, params$Sample_size*params$Group_total/2) 
params$Sub_xy <- ListTwins <- list()
for (Twins in 1:( params$Sample_size*params$Group_total/2) ) {
  ListTwins[[Twins]] <- c(2*Twins-1, 2*Twins)
}
params$Sub_xy <- ListTwins
corr_Sub = diag(params$Sample_size*params$Group_total) 

for (i in 1:length(params$Sub_corr)) {
corr_Sub[params$Sub_xy[[i]][1], params$Sub_xy[[i]][2]] = params$Sub_corr[i]^2
corr_Sub[params$Sub_xy[[i]][2], params$Sub_xy[[i]][1]] = params$Sub_corr[i]^2
}
CovHeatPlot(corr_Sub[1:params$Sample_size, 1:params$Sample_size])
CovHeatPlot(corr_Sub[1:10, 1:10])
sigma_Group = c(1:params$Group_total)

corr_Rep = fVarAR1(0.6, params$Time_points)
sigma_Rep = seq(1, 2, length=params$Time_points)

SimCov_TwinsAR <- CovMarginal_sim(corr_Sub, sigma_Group, corr_Rep, sigma_Rep)

sigma_RE <- params$sigma_RE
SimCov <- CovMarginal_sim(corr_Sub, sigma_Group, corr_Rep, sigma_Rep, sigma_RE) 

DT::datatable(SimCov[1:params$Sample_size, 1:params$Sample_size])

CovHeatPlot(SimCov[1:params$Sample_size, 1:params$Sample_size])
CovHeatPlot(SimCov_TwinsAR[1:params$Sample_size, 1:params$Sample_size])

set.seed(123) ## 
sim_DGEList <- wrap_sim_corr(dat, 
                             params$N_Genes,
                             params$eps_SignalGenes, 
                             SingleGroup = TRUE,
                             coef.beta1 = TRT_effect,
                             coef.beta2 = Time_effect, 
                             CovMarginal = SimCov,
                             act.cor = NULL)

str(sim_DGEList)

set.seed(1) ## 
sim_DGEList <- wrap_sim_corr(dat, 
                             params$N_Genes,
                             params$eps_SignalGenes, 
                             SingleGroup = TRUE,
                             coef.beta1 = TRT_effect,
                             coef.beta2 = Time_effect, 
                             CovMarginal = SimCov_TwinsAR,
                             act.cor = NULL)

str(sim_DGEList)
```

Design and contrast

```{r, eval=FALSE}
## wide and long data frame
norm_counts_wider <- t(sim_DGEList$norm_counts) %>% as.data.frame() %>%
  rownames_to_column(., var = "SampleID") 
annot_table <- sim_DGEList$pheno_data %>% as.data.frame() %>%
  rownames_to_column(., var = "SampleID") 
norm_counts_wider <- annot_table %>% left_join( norm_counts_wider, by=c("SampleID"))

norm_counts_long <- norm_counts_wider %>% tidyr::pivot_longer(
  cols=!(c("SampleID", "ID", "VISIT", "GROUP")),
  names_to = "GeneName", 
  values_to = "Exprs")
```


```{r Add_Family_Info}
norm_counts_wider <- t(sim_DGEList$norm_counts) %>% as.data.frame() %>%
  rownames_to_column(., var = "SampleID") 
annot_table <- sim_DGEList$pheno_data %>% as.data.frame() %>%
  rownames_to_column(., var = "SampleID") %>%
  mutate(FAMILY = ID)
ListFamily <- lapply(1:length(params$Sub_xy), function(n){paste0("P", params$Sub_xy[[n]])} ) 
for (FamilyNum in 1:length(ListFamily)) {
  annot_table <- annot_table %>% mutate(
    FAMILY = ifelse(ID %in% ListFamily[[FamilyNum]], paste0("Family_", FamilyNum), FAMILY)
  )
}

norm_counts_wider <- annot_table %>% left_join( norm_counts_wider, by=c("SampleID"))

norm_counts_long <- norm_counts_wider %>% tidyr::pivot_longer(
  cols=!(c("SampleID", "ID", "FAMILY",  "VISIT", "GROUP")),
  names_to = "GeneName", 
  values_to = "Exprs")

```


Write the simulated twins data to file:

```{r, eval=FALSE}
## save simulated data to file (converge)
# filename = paste0(getwd(), "/data/Simulation_TwinsAR.csv")
# sim_DGE_long_Twins <- norm_counts_long
# write.csv(sim_DGE_long_Twins, file = filename)

# Simulation_Twins <- read_csv("data/Simulation_Twins.csv")
# norm_counts_long <- Simulation_Twins

filename = paste0(getwd(), "/data/Simulation_TwinsCS_Seed123.csv")
sim_DGE_long_TwinsCS <- norm_counts_long
write.csv(sim_DGE_long_TwinsCS, file = filename)

filename = paste0(getwd(), "/data/Simulation_TwinsAR_Seed1.csv")
sim_DGE_long_TwinsAR <- norm_counts_long
write.csv(sim_DGE_long_TwinsAR, file = filename)
```


```{r plot null and signal genes}
### GENE name
plot_prof(norm_counts_long %>% filter(GeneName %in% c("GENE_1", "GENE_50")))
# DT::datatable(norm_counts_long)
```



```{r}
sessionInfo()
```

